---
title: "scRNA-seq analysis of ex vivo macrophages"
output: html_notebook
---

# Load packages

```{r}
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
options(Seurat.object.assay.version = "v3")
```
# Load and format datasets
```{r}
data_dir <- "~/Oxford/Results/scRNA/"
```

```{r}
dir_10x <- list.dirs(path = paste0(data_dir, "Counts/"), full.names = T, recursive = F) %>%
  paste0("/outs/filtered_feature_bc_matrix/")

data <- lapply(dir_10x, Read10X)
names(data) <- list.dirs(path = paste0(data_dir, "Counts/"), full.names = F, recursive = F)

MP.list <- lapply(data, CreateSeuratObject)

## annotate seurat objects 
id <- c(paste("buffer", 1:4, sep = "_"), paste("oxldl", 1:4, sep = "_") )
for( i in 1:8){
  MP.list[[i]]@meta.data$orig.ident <- id[i]
}
```

# Data integration using the Seurat's CCA pipeline and SCTransform

## scTransfrom

```{r}
## score cell cycle genes
MP.list <- lapply(MP.list, CellCycleScoring, s.features = cc.genes$s.genes, g2m.features = cc.genes$g2m.genes, set.ident = F)

MP.list <- lapply(MP.list, FUN = function(object){
  object$CC.Difference <- object$S.Score - object$G2M.Score
  return(object)
})

```
```{r}
#regress out cell cycle differences and MT%

MP.list <- lapply(MP.list, PercentageFeatureSet, pattern = "^MT-", col.name = "percent.mt")
## filter out cells with mt% higher than 20%
MP.list <- lapply(MP.list, subset, subset=(percent.mt <= 20))

MP.list <- lapply(MP.list, SCTransform, assay = "RNA", verbose=F, vars.to.regress = c("CC.Difference", "percent.mt"), variable.features.n = 3000)
```

## CCA integration
```{r }
features <- SelectIntegrationFeatures(object.list = MP.list, nfeatures = 3000)
MP.list <- PrepSCTIntegration(object.list = MP.list, anchor.features = features)

anchors <- FindIntegrationAnchors(object.list = MP.list, 
                                  normalization.method = "SCT", anchor.features = features,
                                  dims = 1:30, reduction = "cca", k.anchor = 5)

MP <- IntegrateData(anchorset = anchors, normalization.method = "SCT", new.assay.name = "integrated", k.weight=100, dims = 1:30)
```


# Clustering parameters optimization - stability test
```{r }
DefaultAssay(MP) <- "integrated"
MP <- RunPCA(MP, verbose = FALSE)
ElbowPlot(MP, ndims = 40)
```
## Optimizing clustering parameters with scclusteval
```{r}
subsample_idents<- readRDS(paste0(data_dir, "/scclusteval/gather_subsample.rds"))

fullsample_idents<- readRDS(paste0(data_dir, "/scclusteval/gather_full_sample.rds"))
```
```{r}
## how many clusters for each different comibination of parameter set?
fullsample_idents %>%
  mutate(cluster_num = purrr::map_dbl(original_ident_full, ~length(unique(.x)))) %>% 
  mutate(id=row_number()) %>% 
  select(id, pc, resolution, k_param, cluster_num)
```

## assign stable clusters based on n=50 of sub-sampling
```{r}
subsample_idents_list<- subsample_idents %>%
  group_by(pc, resolution, k_param) %>%
  nest()

stable_clusters<- subsample_idents_list %>%
  mutate(stable_cluster = map(data, ~ scclusteval::AssignStableCluster(.x$original_ident,
                                                          .x$recluster_ident,
                                                          jaccard_cutoff = 0.8,
                                                          method = "jaccard_percent",
                                                          percent_cutoff = 0.8)))

```
## get the parameter sets with 100% of cells in stable clusters
```{r}
n_stable_clusters <- stable_clusters$stable_cluster %>% lapply(FUN=function(x){return(x$number_of_stable_cluster)}) %>% unlist

df <- fullsample_idents %>%
  mutate(cluster_num = purrr::map_dbl(original_ident_full, ~length(unique(.x)))) %>% 
  mutate(id=row_number()) %>% 
  select(id, pc, resolution, k_param, cluster_num)
df <- df[df$cluster_num==n_stable_clusters,]

df
```


```{r}
p1 <- scclusteval::ParameterSetScatterPlot(stable_clusters = stable_clusters,
                        fullsample_idents = fullsample_idents,
                        x_var = "k_param",
                        y_var = "number",
                        facet_rows = "resolution",
                        facet_cols = "pc") + 
   ggtitle("the number of  clusters identified")

p2 <- scclusteval::ParameterSetScatterPlot(stable_clusters = stable_clusters,
                        fullsample_idents = fullsample_idents,
                        x_var = "k_param",
                        y_var = "percentage",
                        facet_rows = "resolution",
                        facet_cols = "pc") +
  ggtitle("percentage of cells in stable clusters")
```

# Clustering n_PCs = 25, k.param=40, resolution=0.4

```{r}
MP <- RunUMAP(MP, reduction = "pca", dims = 1:25)
```
```{r}
MP <- FindNeighbors(MP, reduction = "pca", dims = 1:25, k.param = 40)
MP <- FindClusters(MP, resolution = 0.4)
```
```{r}
MP$RNA_clusters <- MP$seurat_clusters

DimPlot(MP, reduction = "umap", group.by = "RNA_clusters", label = T, repel = T)
```


## Visualization by donor and by treatment
```{r}
MP$batch <- MP$orig.ident
MP$treatment <- MP$orig.ident %>% 
  gsub(pattern="^([a-z]+)_[1-4]", replacement="\\1")
MP$donor <- paste0("donor_",  gsub(x = MP$orig.ident, pattern="^[a-z]+_([1-4])", replacement="\\1") )
```

```{r}
p1 <- DimPlot(MP, reduction = "umap", group.by = "treatment")

p2 <- DimPlot(MP, reduction = "umap", group.by = "donor")
p1 + p2
```

## Annotation

### Wilcoxon test
```{r}
Rcpp::sourceCpp("~/Oxford/Code/R/functions/fast_wilcox.cpp")
source("~/Oxford/Code/R/functions/Fast_Wilcoxon.R")


DefaultAssay(MP) <- "RNA"
## log1p normalize the data
MP <- NormalizeData(MP)
markers <- wilcoxauc(MP, group_by = "RNA_clusters")
```

```{r}
markers %>% filter(logFC>0) %>% group_by(group) %>% top_n(n=5, wt=-padj)
```
```{r}
MP$subtype <- plyr::revalue(MP$RNA_clusters, c("0" = "FUCA1hi MΦ-1",
                                               "1" = "PCSK6hi MΦ",
                                               "2" = "MHChi MΦ",
                                               "3" = "Lipid-handling MΦ",
                                               "4" = "Cycling MΦ-1",
                                               "5" = "Cycling MΦ-2",
                                               "6" = "Damaged cells (excluded)",
                                               "7" = "T-cells (excluded)",
                                               "8" = "M1-like MΦ"))
```

```{r}
saveRDS(MP, paste0(data_dir,"MP.Nov2023.rds"))
```

```{r}
MP_oxldl <- readRDS(paste0(data_dir,"MP.Nov2023.rds")) %>% 
  subset(RNA_clusters %in% c(0:5,8))
```

